 

### 2021-07-27 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

Leaf thermal tolerances are more closely adapted to microclimate than macroclimate [@feeley_thermal_2020; @slotPhotosyntheticPlasticityTropical2021]. Chinquapin oak (*Quercus muehlenbergii*) growing in xeric, sunnier conditions showed higher $T_{50}$ by $2{^\circ}C$ than bur oak (*Quercus macrocarpa*) growing in shaded mesic conditions [@hamerlynckPhotosyntheticStomatalResponses1996a]. Along a vertical gradient within canopy, *Acacia papyrocarpa* (Benth.) showed greater $T_{50}$ and higher composite climate stress in the lower north-facing canopy than other canopy positions, correlating with low wind-speed, greater $T_{air}$ and longer leaf thermal time constants [@curtis_intracanopy_2019]. Additionally, thermal tolerances are explained more by leaf traits than phylogeny [@feeley_thermal_2020; @slotPhotosyntheticPlasticityTropical2021]. $T_{crit}$ was found to be greater among tropical species with large leaves that had slow heating and cooling time, whereas $T_{50}$ was greater among species with greater LMA [@slotLeafHeatTolerance2020]. Across species sun leaves that experienced higher maximum temperatures show higher photosynthetic heat tolerance [@perezPhotosyntheticHeatTolerances2020]. However, considering sun and shade leaf differences within canopy (Table 1) $T_{50}$ varies modestly, being slightly lower for shade than sun leaves for two of three tree species in Panama [@slotPhotosyntheticHeatTolerance2019a]. Based on this evidence, we would expect leaves exposed to greater heat stress within the canopy to have higher heat tolerance, but further research is required to fully understand the relationship between heat tolerance, light exposure and microclimates.

@@

**Similar to photosynthesis, respiration tends to be higher in canopy sun-leaves [@chenLeafEconomicsSpectrum2020], but temperature sensitivity of respiration--$Q_{10}$--shows no definite trend along the vertical gradient [@bolstad_foliar_1999; Weerasinghe et al.2014].**

Temperature sensitivity of respiration ($Q_{10}$) along a within canopy gradient remains uncertain where $Q_{10}$ can be constant within vertical profiles and in seedling sun vs. shade leaves [@@carterExperimentalWarmingTropical; @weerasingheCanopyPositionAffects2014; @bolstad_foliar_1999; @zaragoza-castellsDoesGrowthIrradiance2007; @zaragoza-castellsClimatedependentVariationsLeaf2008; @xuSeasonalVariationTemperature2006], and greater in upper-canopy leaves [@turnbullScalingFoliarRespiration2003; @harleyEffectsLightTemperature1996]. Additionally, lower canopy leaf respiration has also been reported to be more sensitive to changes in $T_{air}$ than upper canopy [@griffinCanopyPositionAffects2002]. Recent observations suggest that tropical understory leaves down-regulate $Q_{10}$ with rising temperature [@carterExperimentalWarmingTropical]; this may vary based on leaf age and leaf traits, as in evergreen species, $Q_{10}$ has shown to be greater in younger than older leaves [@zhouLeafageEffectsTemperature2015].

>>>>

Leaf thermal tolerances are more closely adapted to microclimate than macroclimate [@feeley_thermal_2020; @slotPhotosyntheticPlasticityTropical2021]. Chinquapin oak (*Quercus muehlenbergii*) growing in xeric, sunnier conditions showed higher $T_{50}$ by $2{^\circ}C$ than bur oak (*Quercus macrocarpa*) growing in shaded mesic conditions [@hamerlynckPhotosyntheticStomatalResponses1996a]. Along a vertical gradient within the canopy, *Acacia papyrocarpa* (Benth.) showed greater $T_{50}$ and higher composite climate stress in the lower, north-facing canopy than other canopy positions, correlating with low wind-speed, greater radiation and $T_{air}$, and lower rates of heat dissapation, as indicated by longer leaf thermal time constants [@curtis_intracanopy_2019]. Additionally, thermal tolerances are explained more by leaf traits than phylogeny [@feeley_thermal_2020; @slotPhotosyntheticPlasticityTropical2021]. $T_{crit}$ was found to be greater among tropical species with large leaves that had slow heating and cooling time, whereas $T_{50}$ was greater among species with greater LMA [@slotPhotosyntheticPlasticityTropical2021]. Across species sun leaves that experienced higher maximum temperatures show higher photosynthetic heat tolerance [@perezPhotosyntheticHeatTolerances2020]. However, considering sun and shade leaf differences within canopy (Table 1), $T_{50}$ varies modestly, being slightly lower for shade than sun leaves for two of three tree species in Panama [@slotPhotosyntheticHeatTolerance2019a]. We would expect leaves exposed to greater heat stress within the canopy to have higher heat tolerance, but further research is required to fully understand the relationship between heat tolerance, light exposure and microclimates.

@@

**Similar to photosynthesis, respiration tends to be higher in canopy sun-leaves [Table 2, @chenLeafEconomicsSpectrum2020], but its temperature sensitivity ($Q_{10}$) shows no definite trend along the vertical gradient [@bolstad_foliar_1999; Weerasinghe et al.2014].** Specifically, the $Q_{10}$ of respiration can be constant within vertical profiles and in seedling sun vs. shade leaves [@@carterExperimentalWarmingTropical; @weerasingheCanopyPositionAffects2014; @bolstad_foliar_1999; @zaragoza-castellsDoesGrowthIrradiance2007; @zaragoza-castellsClimatedependentVariationsLeaf2008; @xuSeasonalVariationTemperature2006], greater in upper-canopy leaves [@turnbullScalingFoliarRespiration2003; @harleyEffectsLightTemperature1996], or greater in the lower canopy [@griffinCanopyPositionAffects2002]. Recent observations suggest that tropical understory leaves down-regulate $Q_{10}$ with rising temperature [@carterExperimentalWarmingTropical]; this may vary based on leaf age and leaf traits, as in evergreen species, $Q_{10}$ has shown to be greater in younger than older leaves [@zhouLeafageEffectsTemperature2015].
